---
title: "qPCR Transporter Figures"
author: "C-T Berezin"
date: "10/30/2021"
output:
  pdf_document: default
---

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(ggthemes)
library(forcats)
library(ggpubr)
#library(writexl)
library(emmeans)
library(lme4)
library(car)
library(lmerTest)
library(svglite)
```

# Comparing chronic saline/morphine for Bcrp (rerun of samples for P-gp from Bergum 2022)

## Reading in & tidying the data

```{r}
transporters <- read.csv("../../data/qRT-PCR/07222022-transporter-gene-study.csv", fileEncoding = 'UTF-8-BOM')

transporters <- transporters %>% filter(Sample != "calib")
  
transporters <- transporters %>% mutate(Tissue = str_sub(Sample, start=-3L, end=-1L),
                                        Treatment = str_extract(Sample, "[:alpha:]+(?=[:digit:])"),
                                        Sample = str_extract(Sample, "[:alnum:]+(?=-)"))

transporters <- transporters %>% mutate(Treatment = factor(Treatment, c("S", "M", "Flo", "Fhi", "C")),
                                        Tissue = factor(Tissue, c("ret", "hyp")))

head(transporters)

transporters_long <- transporters %>% pivot_longer(cols=c("Pgp.RGE", "Bcrp.RGE"), names_to = "gene", values_to="rge")

transporters_long <- transporters_long %>% mutate(gene = str_extract(gene, "[:alpha:]+(?=.)"))
transporters_long <- transporters_long %>% mutate(gene = factor(as.factor(gene), c("Pgp", "Bcrp")))


head(transporters_long)
```

## Stats on Bcrp only (chronic vs saline)

```{r}
#bcrp only
chronic <- transporters %>% filter(Treatment %in% c("C", "S"))
chronic

c_bcrp_lm <- lmer(Bcrp.RGE ~ Tissue * Treatment + (1|Sample), data=chronic)
plot(c_bcrp_lm, type=c("p","smooth"), col.line=1)
lattice::qqmath(c_bcrp_lm)

c_bcrp_log2_lm <- lmer(log2(Bcrp.RGE) ~ Tissue * Treatment + (1|Sample), data=chronic)
plot(c_bcrp_log2_lm, type=c("p","smooth"), col.line=1)
lattice::qqmath(c_bcrp_log2_lm)

shapiro.test(chronic$Bcrp.RGE)
shapiro.test(log2(chronic$Bcrp.RGE))

anova(c_bcrp_log2_lm)

emmeans::emmeans(c_bcrp_log2_lm, pairwise ~ Treatment | Tissue)$contrasts
emmeans::emmeans(c_bcrp_log2_lm, pairwise ~ Tissue | Treatment)$contrasts


c_bcrp_sumstats <- chronic %>% group_by(Treatment, Tissue) %>% summarise(
  n = n(),
  mean = mean(Bcrp.RGE),
  sd = sd(Bcrp.RGE),
  se = sd/sqrt(n),
  log2_mean = mean(log2(Bcrp.RGE)),
  log2_sd = sd(log2(Bcrp.RGE)),
  log2_se = log2_sd/sqrt(n)
)
c_bcrp_sumstats
```

## Visualizing chronic vs saline Bcrp

```{r}
#visualizing tissue effect per treatment
c_bcrp_tiss_plot <- ggplot() +
  stat_summary(data=chronic, aes(x=Treatment, y=Bcrp.RGE, fill=Tissue), fun='mean',
               geom="bar", width=0.75, position=position_dodge(width=0.75)) +
  geom_errorbar(data=c_bcrp_sumstats, aes(x=Treatment, ymin=mean-se,
                                           ymax=mean+se, fill=Tissue),
                width=0.25, position=position_dodge(width=0.75)) +
  geom_point(data=chronic, aes(x=Treatment, y=Bcrp.RGE, fill=Tissue),
             position=position_dodge(width=0.75)) +
  #facet_grid(~Tissue, scales="free_y") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  theme_bw(base_size = 12) %+replace%
    theme(legend.title = element_blank()) +
  scale_fill_manual(values=c("dodgerblue", "goldenrod1"),
                    labels=c("Retina", "Brain")) +
  scale_x_discrete(labels=c("S" = "Saline", "C" = "Morphine")) +
  xlab("") +
  ylab(expression(RGE[Bcrp]))

c_bcrp_tiss_plot
#ggsave(filename="../../figures/qRT-PCR/c_bcrp_tiss_plot.png", plot=c_bcrp_tiss_plot, height=4, width=4)
#ggsave(filename="../../figures/qRT-PCR/c_bcrp_tiss_plot.svg", plot=c_bcrp_tiss_plot, height=4, width=4)

c_bcrp_tiss_log_plot <- ggplot() +
  stat_summary(data=chronic, aes(x=Treatment, y=log2(Bcrp.RGE), fill=Tissue),
               fun = 'mean', geom="bar", width=0.75, position="dodge") +
  geom_errorbar(data=c_bcrp_sumstats, aes(x=Treatment, ymin=log2_mean-log2_se,
                                         ymax=log2_mean+log2_se, fill=Tissue),
                width=0.25, position=position_dodge(width=0.75)) +
  geom_point(data=chronic, aes(x=Treatment, y=log2(Bcrp.RGE), fill=Tissue),
             stat="identity", position=position_dodge(width=0.75)) +
  #facet_grid(~Tissue, scales="free_y") +
  #scale_y_continuous(expand = expansion(mult = c(0.1, 0.2))) +
  ylim(-0.6, 2.5) +
  #scale_fill_manual(values=c("maroon1", "springgreen3")) +
  theme_bw(base_size = 12) %+replace%
    theme(legend.title = element_blank(),
          legend.box.spacing = unit(1, "pt"),
          axis.title.x=element_blank(),
          plot.title = element_text(hjust = 0.5, vjust=1.5)
          ) +
  scale_fill_manual(values=c("dodgerblue", "goldenrod1"),
                    labels=c("Retina", "Brain")) +
  scale_x_discrete(labels=c("S" = "Saline", "C" = "Morphine")) +
  xlab("") +
  ylab(expression(log[2](RGE)))

c_bcrp_tiss_log_plot
#ggsave(filename="../../figures/qRT-PCR/c_bcrp_tiss_log_plot.png", plot=c_bcrp_tiss_log_plot, height=4, width=4)
#ggsave(filename="../../figures/qRT-PCR/c_bcrp_tiss_log_plot.svg", plot=c_bcrp_tiss_log_plot, height=4, width=4)


#visualizing treatment effect per tissue
c_bcrp_trt_plot <- ggplot() +
  stat_summary(data=chronic, aes(x=Tissue, y=Bcrp.RGE, fill=Treatment),
               fun = 'mean', geom="bar", width=0.75, position="dodge") +
  geom_errorbar(data=c_bcrp_sumstats, aes(x=Tissue, ymin=mean-se,
                                           ymax=mean+se, fill=Treatment),
                width=0.25, position=position_dodge(width=0.75)) +
  geom_point(data=chronic, aes(x=Tissue, y=Bcrp.RGE, fill=Treatment),
             stat="identity", position=position_dodge(width=0.75)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  theme_bw(base_size = 12) %+replace%
    theme(legend.title = element_blank()) +
  scale_fill_manual(values=c("gray70", "red4"),
                    labels=c("Saline", "Morphine")) +
  scale_x_discrete(labels=c("ret" = "Retina", "hyp" = "Brain")) +
  xlab("") +
  ylab(expression(RGE[Bcrp]))

c_bcrp_trt_plot
#ggsave(filename="../../figures/qRT-PCR/c_bcrp_trt_plot.png", plot=c_bcrp_trt_plot, height=4, width=4)
#ggsave(filename="../../figures/qRT-PCR/c_bcrp_trt_plot.svg", plot=c_bcrp_trt_plot, height=4, width=4)

c_bcrp_trt_log_plot <- ggplot() +
  stat_summary(data=chronic, aes(x=Tissue, y=log2(Bcrp.RGE), fill=Treatment),
               fun = 'mean', geom="bar", width=0.75, position="dodge") +
  geom_errorbar(data=c_bcrp_sumstats, aes(x=Tissue, ymin=log2_mean-log2_se,
                                         ymax=log2_mean+log2_se, fill=Treatment),
                width=0.25, position=position_dodge(width=0.75)) +
  geom_point(data=chronic, aes(x=Tissue, y=log2(Bcrp.RGE), fill=Treatment),
             stat="identity", position=position_dodge(width=0.75)) +
  #scale_y_continuous(expand = expansion(mult = c(0.1, 0.25))) +
  ylim(-0.6, 2.5) +
  theme_bw(base_size = 12) %+replace%
    theme(legend.title = element_blank(),
          legend.box.spacing = unit(1, "pt"),
          axis.title.x=element_blank(),
          plot.title = element_text(hjust = 0.5, vjust=1.5)) +
  scale_fill_manual(values=c("gray70", "red4"),
                    labels=c("Saline", "Morphine")) +
  scale_x_discrete(labels=c("ret" = "Retina", "hyp" = "Brain")) +
  xlab("") +
  ylab(expression(log[2](RGE)))

c_bcrp_trt_log_plot

#ggsave(filename="../../figures/qRT-PCR/c_bcrp_trt_log_plot.png", plot=c_bcrp_trt_log_plot, height=4, width=4)
#ggsave(filename="../../figures/qRT-PCR/c_bcrp_trt_log_plot.svg", plot=c_bcrp_trt_log_plot, height=4, width=4)
```

### Chronic morphine Bcrp test WITH P-gp in stats

```{r}
#bcrp & pgp together
chronic_long <- transporters_long %>% filter(Treatment %in% c("C", "S"))
chronic_long

chronic_lm <- lmer(rge ~ gene * Tissue * Treatment + (1|Sample), data=chronic_long)
plot(chronic_lm, type=c("p","smooth"), col.line=1)
lattice::qqmath(chronic_lm)

chronic_log2_lm <- lmer(log2(rge) ~ gene * Tissue * Treatment + (1|Sample), data=chronic_long)
plot(chronic_log2_lm, type=c("p","smooth"), col.line=1)
lattice::qqmath(chronic_log2_lm)

shapiro.test(chronic_long$rge)
shapiro.test(log2(chronic_long$rge))


anova(chronic_log2_lm)

emmeans::emmeans(chronic_log2_lm, pairwise ~ Treatment | Tissue, by="gene")$contrasts
emmeans::emmeans(chronic_log2_lm, pairwise ~ Tissue | Treatment, by="gene")$contrasts
emmeans::emmeans(chronic_log2_lm, pairwise ~ gene | Treatment, by="Tissue")$contrasts

chronic_sumstats <- chronic_long %>% group_by(Treatment, Tissue, gene) %>% summarise(
  n = n(),
  mean = mean(rge),
  sd = sd(rge),
  se = sd/sqrt(n),
  log2_mean = mean(log2(rge)),
  log2_sd = sd(log2(rge)),
  log2_se = log2_sd/sqrt(n)
)
chronic_sumstats
```

```{r}
chronic_plot <- ggplot() +
  stat_summary(data=chronic_long, aes(x=Tissue, y=rge, fill=Treatment),
               fun = 'mean', geom="bar", width=0.75, position="dodge") +
  geom_errorbar(data=chronic_sumstats, aes(x=Tissue, ymin=mean-sd,
                                           ymax=mean+sd, fill=Treatment),
                width=0.25, position=position_dodge(width=0.75)) +
  geom_point(data=chronic_long, aes(x=Tissue, y=rge, fill=Treatment),
             stat="identity", position=position_dodge(width=0.75)) +
  facet_grid(~gene, scales="free_y") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  #scale_fill_manual(values=c("maroon1", "springgreen3")) +
  theme_bw()

chronic_plot

#ggsave(filename="../../figures/qRT-PCR/chronic_trt_plot.svg", plot=chronic_plot, height=4, width=6)


chronic_log_plot <- ggplot() +
  stat_summary(data=chronic_long, aes(x=Tissue, y=log2(rge), fill=Treatment),
               fun = 'mean', geom="bar", width=0.75, position="dodge") +
  geom_errorbar(data=chronic_sumstats, aes(x=Tissue, ymin=log2_mean-log2_sd,
                                         ymax=log2_mean+log2_sd, fill=Treatment),
                width=0.25, position=position_dodge(width=0.75)) +
  geom_point(data=chronic_long, aes(x=Tissue, y=log2(rge), fill=Treatment),
             stat="identity", position=position_dodge(width=0.75)) +
  facet_grid(~gene, scales="free_y") +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.25))) +
  #scale_fill_manual(values=c("maroon1", "springgreen3")) +
  theme_bw()

chronic_log_plot

#ggsave(filename="../../figures/qRT-PCR/chronic_trt_log.svg", plot=chronic_log_plot, height=4, width=6)
```


### RERUN STATS WITH LOG2 ON DATA ORIGINALLY PUBLISHED IN BERGUM 2022 (p-gp chronic morphine/saline)

```{r}
pgp <- read.csv("../../data/qRT-PCR/pgp-Bergum2022a.csv", fileEncoding = 'UTF-8-BOM')
pgp <- pgp %>% mutate(Tissue = fct_relevel(Tissue, c("retina", "hypothalamus")),
                      Treatment = fct_relevel(as.factor(Treatment), c("saline", "morphine")))
#removing the outlier
pgp <- pgp %>% filter(RelativeGE<3.5)
pgp

shapiro.test(pgp$RelativeGE)
shapiro.test(sqrt(pgp$RelativeGE))
#normal
shapiro.test(log(pgp$RelativeGE))
shapiro.test(log2(pgp$RelativeGE))

pgp <- pgp %>% mutate(logRGE = log(RelativeGE))
pgp <- pgp %>% mutate(log2RGE = log2(RelativeGE))

car::leveneTest(logRGE ~ Tissue * Treatment, pgp)
pgp_log_lm <- lm(logRGE ~ Treatment * Tissue, pgp)
plot(pgp_log_lm, which=c(1,2))
anova(pgp_log_lm)
emmeans::emmeans(pgp_log_lm, pairwise ~ Treatment | Tissue)$contrasts
emmeans::emmeans(pgp_log_lm, pairwise ~ Tissue | Treatment)$contrasts

pgp_log_lmer <- lmer(log(RelativeGE) ~ Tissue * Treatment + (1|Sample), data =pgp)
plot(pgp_log_lmer, type=c("p","smooth"), col.line=1)
lattice::qqmath(pgp_log_lmer)
anova(pgp_log_lmer)
emmeans::emmeans(pgp_log_lmer, pairwise ~ Treatment | Tissue)
emmeans::emmeans(pgp_log_lmer, pairwise ~ Tissue | Treatment)

pgp_log2_lmer <- lmer(log2(RelativeGE) ~ Tissue * Treatment + (1|Sample), data =pgp)
plot(pgp_log2_lmer, type=c("p","smooth"), col.line=1)
lattice::qqmath(pgp_log2_lmer)
anova(pgp_log2_lmer)
emmeans::emmeans(pgp_log2_lmer, pairwise ~ Treatment | Tissue)
emmeans::emmeans(pgp_log2_lmer, pairwise ~ Tissue | Treatment)
```

```{r}
pgp_log2_tiss_plot <- pgp %>% ggplot(aes(x=Treatment, y=log2RGE, fill=Tissue)) +
  stat_summary(fun = 'mean', geom="bar", aes(fill=Tissue),
               width=0.75, position="dodge") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width=0.25,
               position=position_dodge(width=0.75)) +
  geom_point(stat="identity", position=position_dodge(width=0.75)) +
  theme_bw(base_size = 12) %+replace%
  theme(axis.title.x=element_blank(),
        legend.title = element_blank(),
        legend.box.spacing = unit(1, "pt"),
        plot.title = element_text(hjust = 0.5, vjust=1.5)) +
  labs(y="log(Relative Gene Expression)") +
  #scale_y_continuous(expand = expansion(mult = c(0.1, 0.2))) +
  ylim(-0.6, 2) +
  scale_fill_manual(values = c("dodgerblue", "goldenrod1"),
                    labels = c("Retina", "Brain")) +
  scale_x_discrete(labels=c("saline" = "Saline", "morphine" = "Morphine")) +
  ylab(expression(log[2](RGE)))

pgp_log2_tiss_plot
#ggsave(filename = "../figures/pgp_log2_tiss_plot.png", plot = pgp_log2_tiss_plot, height=4, width=4)


pgp_log2_trt_plot <- pgp %>% ggplot(aes(x=Tissue, y=log2RGE, fill=Treatment)) +
  stat_summary(fun = 'mean', geom="bar", aes(fill=Treatment),
               width=0.75, position="dodge") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width=0.25,
               position=position_dodge(width=0.75)) +
  geom_point(stat="identity", position=position_dodge(width=0.75)) +
  theme_bw(base_size = 12) %+replace%
  theme(axis.title.x=element_blank(),
        legend.title = element_blank(),
        legend.box.spacing = unit(1, "pt"),
        plot.title = element_text(hjust = 0.5, vjust=1.5)) +
  labs(y="log(Relative Gene Expression)") +
  #scale_y_continuous(expand = expansion(mult = c(0.1, 0.25))) +
  ylim(-0.6, 2) +
  scale_fill_manual(values = c("gray70", "red4"),
                    labels=c("Saline", "Morphine")) +
  scale_x_discrete(labels=c("retina" = "Retina", "hypothalamus" = "Brain")) +
  ylab(expression(log[2](RGE)))

pgp_log2_trt_plot
#ggsave(filename = "../../figures/pgp_log2_trt_plot.png", plot = pgp_log2_trt_plot, height=4, width=4)
```


# Looking at only the 1 injection samples (males, low e/p females, high e/p females)

## Reading in the new dataset

```{r}
transporters <- read.csv("../../data/qRT-PCR/07222022-transporter-gene-study-1inj-only.csv", fileEncoding = 'UTF-8-BOM')

#transporters <- transporters %>% filter(Sample != "calib")
  
transporters <- transporters %>% mutate(Tissue = str_sub(Sample, start=-3L, end=-1L),
                                        Group = str_extract(Sample, "[:alpha:]+(?=[:digit:])"),
                                        Sample = str_extract(Sample, "[:alnum:]+(?=-)"))

transporters <- transporters %>% mutate(Group = factor(Group, c("M", "Flo", "Fhi")),
                                        Tissue = factor(Tissue, c("ret", "hyp")),
                                        Sample = as.factor(Sample))

transporters_long <- transporters %>% pivot_longer(cols=c("Pgp.RGE", "Bcrp.RGE"), names_to = "gene", values_to="rge")

transporters_long <- transporters_long %>% mutate(gene = str_extract(gene, "[:alpha:]+(?=.)"))
transporters_long <- transporters_long %>% mutate(gene = factor(as.factor(gene), c("Pgp", "Bcrp")))

transporters_long
```

## Stats

```{r}
transporter_lm <- lmer(rge ~ gene * Tissue * Group + (1|Sample), data=transporters_long)
plot(transporter_lm, type=c("p","smooth"), col.line=1)
lattice::qqmath(transporter_lm)

transporter_log2_lm <- lmer(log2(rge) ~ gene * Tissue * Group + (1|Sample), data=transporters_long)
plot(transporter_log2_lm, type=c("p","smooth"), col.line=1)
lattice::qqmath(transporter_log2_lm)

shapiro.test(transporters_long$rge)
shapiro.test(log2(transporters_long$rge))


anova(transporter_log2_lm)

emmeans::emmeans(transporter_log2_lm, pairwise ~ Group | Tissue, by="gene")$contrasts
emmeans::emmeans(transporter_log2_lm, pairwise ~ Tissue | Group, by="gene")$contrasts
emmeans::emmeans(transporter_log2_lm, pairwise ~ gene | Group, by="Tissue")$contrasts
```

```{r}
trans_sumstats <- transporters_long %>% group_by(Group, Tissue, gene) %>% summarise(
  n = n(),
  mean = mean(rge),
  sd = sd(rge),
  se = sd/sqrt(n),
  log2_mean = mean(log2(rge)),
  log2_sd = sd(log2(rge)),
  log2_se = log2_sd/sqrt(n)
)
trans_sumstats
```

## Visualizing hyp vs ret within each treatment/gene

```{r}
gene.labs <- c("P-gp", "Bcrp")
names(gene.labs) <- c("Pgp", "Bcrp")

trans_hyp_ret_plot <- ggplot() +
  stat_summary(data=transporters_long, aes(x=Group, y=rge, fill=Tissue),
               fun = 'mean', geom="bar", width=0.75, position="dodge") +
  geom_errorbar(data=trans_sumstats, aes(x=Group, ymin=mean-se, ymax=mean+se, fill=Tissue),
                width=0.25, position=position_dodge(width=0.75)) +
  geom_point(data=transporters_long, aes(x=Group, y=rge, fill=Tissue),
             stat="identity", position=position_dodge(width=0.75)) +
  facet_wrap(~gene, labeller = labeller(gene = gene.labs), ncol=2) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  theme_bw(base_size = 12) %+replace%
  theme(legend.title = element_blank()) +
  scale_fill_manual(values=c("dodgerblue", "goldenrod1"),
                    labels=c("Retina", "Brain")) +
  scale_x_discrete(labels=c("M" = "Males", "Flo" = "Low E/P\nFemales",
                            "Fhi" = "High E/P\nFemales")) +
  xlab("") +
  ylab("RGE")

trans_hyp_ret_plot
#ggsave(filename="../../figures/qRT-PCR/trans_hyp_ret_1inj.png", plot=trans_hyp_ret_plot, height=4, width=7)
#ggsave(filename="../../figures/qRT-PCR/trans_hyp_ret_1inj.svg", plot=trans_hyp_ret_plot, height=4, width=7)

trans_hyp_ret_log_plot <- ggplot() +
  stat_summary(data=transporters_long, aes(x=Group, y=log2(rge), fill=Tissue),
               fun = 'mean', geom="bar", width=0.75, position="dodge") +
  geom_errorbar(data=trans_sumstats, aes(x=Group, ymin=log2_mean-log2_se,
                                         ymax=log2_mean+log2_se, fill=Tissue),
                width=0.25, position=position_dodge(width=0.75)) +
  geom_point(data=transporters_long, aes(x=Group, y=log2(rge), fill=Tissue),
             stat="identity", position=position_dodge(width=0.75)) +
  facet_wrap(~gene, labeller = labeller(gene = gene.labs), ncol=2) +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.25))) +
  theme_bw(base_size = 12) %+replace%
  theme(legend.title = element_blank()) +
  scale_fill_manual(values=c("dodgerblue", "goldenrod1"),
                    labels=c("Retina", "Brain")) +
  scale_x_discrete(labels=c("M" = "Males", "Flo" = "Low E/P\nFemales",
                            "Fhi" = "High E/P\nFemales")) +
  xlab("") +
  ylab(expression(log[2](RGE)))

trans_hyp_ret_log_plot
#ggsave(filename="../../figures/qRT-PCR/trans_hyp_ret_log_1inj.png", plot=trans_hyp_ret_log_plot, height=4, width=7)
#ggsave(filename="../../figures/qRT-PCR/trans_hyp_ret_log_1inj.svg", plot=trans_hyp_ret_log_plot, height=4, width=7)
```

## Visualizing sex diffs within each tissue/gene

```{r}
trans_trt_plot <- ggplot() +
  stat_summary(data=transporters_long, aes(x=Tissue, y=rge, fill=Group),
               fun = 'mean', geom="bar", width=0.75, position="dodge") +
  geom_errorbar(data=trans_sumstats, aes(x=Tissue, ymin=mean-se, ymax=mean+se, fill=Group),
                width=0.25, position=position_dodge(width=0.75)) +
  geom_point(data=transporters_long, aes(x=Tissue, y=rge, fill=Group),
             stat="identity", position=position_dodge(width=0.75)) +
  facet_wrap(~gene, labeller = labeller(gene = gene.labs), ncol=2) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  theme_bw(base_size = 12) %+replace%
  theme(legend.title = element_blank(),
        legend.spacing.y = unit(0.1, 'cm')) +
  ## next line needed to make previous line work
  guides(fill = guide_legend(byrow = TRUE)) +
  scale_fill_manual(values=c("gray50", "darkolivegreen3", "lightpink2"),
                    labels=c("M" = "Males", "Flo" = "Low E/P\nfemales",
                            "Fhi" = "High E/P\nfemales")) +
  scale_x_discrete(labels=c("ret" = "Retina", "hyp" = "Brain")) +
  xlab("") +
  ylab("RGE")

trans_trt_plot
#ggsave(filename="../../figures/qRT-PCR/trans_trt_1inj.png", plot=trans_trt_plot, height=4, width=7)
#ggsave(filename="../../figures/qRT-PCR/trans_trt_1inj.svg", plot=trans_trt_plot, height=4, width=7)


trans_trt_log_plot <- ggplot() +
  stat_summary(data=transporters_long, aes(x=Tissue, y=log2(rge), fill=Group),
               fun = 'mean', geom="bar", width=0.75, position="dodge") +
  geom_errorbar(data=trans_sumstats, aes(x=Tissue, ymin=log2_mean-log2_se,
                                         ymax=log2_mean+log2_se, fill=Group),
                width=0.25, position=position_dodge(width=0.75)) +
  geom_point(data=transporters_long, aes(x=Tissue, y=log2(rge), fill=Group),
             stat="identity", position=position_dodge(width=0.75)) +
  facet_wrap(~gene, labeller = labeller(gene = gene.labs), ncol=2) +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.25))) +
  theme_bw(base_size = 12) %+replace%
  theme(legend.title = element_blank(),
        legend.spacing.y = unit(0.1, 'cm')) +
  ## next line needed to make previous line work
  guides(fill = guide_legend(byrow = TRUE)) +
  scale_fill_manual(values=c("gray50", "darkolivegreen3", "lightpink2"),
                    labels=c("M" = "Males", "Flo" = "Low E/P\nfemales",
                            "Fhi" = "High E/P\nfemales")) +
  scale_x_discrete(labels=c("ret" = "Retina", "hyp" = "Brain")) +
  xlab("") +
  ylab(expression(log[2](RGE)))

trans_trt_log_plot
#ggsave(filename="../../figures/qRT-PCR/trans_trt_log_1inj.png", plot=trans_trt_log_plot, height=4, width=7)
#ggsave(filename="../../figures/qRT-PCR/trans_trt_log_1inj.svg", plot=trans_trt_log_plot, height=4, width=7)
```

## Visualizing P-gp vs Bcrp differences

```{r}
group.labs <- c("Males", "Low E/P Females", "High E/P Females")
names(group.labs) <- c("M", "Flo", "Fhi")

trans_g2g_plot <- ggplot() +
  stat_summary(data=transporters_long, aes(x=Tissue, y=rge, fill=gene),
               fun = 'mean', geom="bar", width=0.75, position="dodge") +
  geom_errorbar(data=trans_sumstats, aes(x=Tissue, ymin=mean-se,
                                         ymax=mean+se, fill=gene),
                width=0.25, position=position_dodge(width=0.75)) +
  geom_point(data=transporters_long, aes(x=Tissue, y=rge, fill=gene),
             position=position_dodge(width=0.75)) +
  facet_grid(~Group, labeller = labeller(Group = group.labs)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  scale_fill_manual(values=c("darkorchid1", "orange1"),
                    labels = c("P-gp", "Bcrp")) +
  theme_bw(base_size = 12) %+replace%
  theme(legend.title = element_blank()) +
  xlab("") +
  ylab("RGE") +
  scale_x_discrete(labels=c("ret" = "Retina", "hyp" = "Brain"))

trans_g2g_plot
#ggsave(filename="../../figures/qRT-PCR/trans_g2g_1inj.png", plot=trans_g2g_plot, height=4, width=7)
#ggsave(filename="../../figures/qRT-PCR/trans_g2g_1inj.svg", plot=trans_g2g_plot, height=4, width=7)


trans_g2g_log_plot <- ggplot() +
  stat_summary(data=transporters_long, aes(x=Tissue, y=log2(rge), fill=gene),
               fun = 'mean', geom="bar", width=0.75, position="dodge") +
  geom_errorbar(data=trans_sumstats, aes(x=Tissue, ymin=log2_mean-log2_se,
                                         ymax=log2_mean+log2_se, fill=gene),
                width=0.25, position=position_dodge(width=0.75)) +
  geom_point(data=transporters_long, aes(x=Tissue, y=log2(rge), fill=gene),
             position=position_dodge(width=0.75)) +
  facet_wrap(~Group, labeller = labeller(Group = group.labs)) +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.25))) +
  scale_fill_manual(values=c("darkorchid1", "orange1"),
                    labels = c("P-gp", "Bcrp")) +
  theme_bw(base_size = 12) %+replace%
  theme(legend.title = element_blank()) +
  xlab("") +
  ylab(expression(log[2](RGE))) +
  scale_x_discrete(labels=c("ret" = "Retina", "hyp" = "Brain"))

trans_g2g_log_plot
#ggsave(filename="../../figures/qRT-PCR/trans_g2g_log_1inj.png", plot=trans_g2g_log_plot, height=4, width=7)
#ggsave(filename="../../figures/qRT-PCR/trans_g2g_log_1inj.svg", plot=trans_g2g_log_plot, height=4, width=7)

```




# To bottom: Full analysis (all 5 groups - M chronic saline, M chronic morphine,
#M acute morphine, Flo & Fhi acute morphine) - not used!!


```{r}
transporter_lm <- lmer(rge ~ gene * Tissue * Group + (1|Sample), data=transporters_long)
plot(transporter_lm, type=c("p","smooth"), col.line=1)
lattice::qqmath(transporter_lm)

transporter_log2_lm <- lmer(log2(rge) ~ gene * Tissue * Group + (1|Sample), data=transporters_long)
plot(transporter_log2_lm, type=c("p","smooth"), col.line=1)
lattice::qqmath(transporter_log2_lm)

shapiro.test(transporters_long$rge)
shapiro.test(log2(transporters_long$rge))


anova(transporter_log2_lm)

emmeans::emmeans(transporter_log2_lm, pairwise ~ Group | Tissue, by="gene")$contrasts
emmeans::emmeans(transporter_log2_lm, pairwise ~ Tissue | Group, by="gene")$contrasts
emmeans::emmeans(transporter_log2_lm, pairwise ~ gene | Group, by="Tissue")$contrasts
```

```{r}
trans_sumstats <- transporters_long %>% group_by(Group, Tissue, gene) %>% summarise(
  n = n(),
  mean = mean(rge),
  sd = sd(rge),
  log2_mean = mean(log2(rge)),
  log2_sd = sd(log2(rge))
)
trans_sumstats
```

### Comparing hyp vs ret within each treatment/gene

```{r}
trans_hyp_ret_plot <- ggplot() +
  stat_summary(data=transporters_long, aes(x=Group, y=rge, fill=Tissue),
               fun = 'mean', geom="bar", width=0.75, position="dodge") +
  geom_errorbar(data=trans_sumstats, aes(x=Group, ymin=mean-sd, ymax=mean+sd, fill=Tissue),
                width=0.2, position=position_dodge(width=0.75)) +
  geom_point(data=transporters_long, aes(x=Group, y=rge, fill=Tissue),
             stat="identity", position=position_dodge(width=0.75)) +
  facet_wrap(~gene, scales="free_y", ncol=1) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  #scale_fill_manual(values=c("maroon1", "springgreen3")) +
  theme_bw()

trans_hyp_ret_plot

#ggsave(filename="../../figures/qRT-PCR/trans_hyp_ret.png", plot=trans_hyp_ret_plot, height=4, width=6)
#ggsave(filename="../../figures/qRT-PCR/trans_hyp_ret.svg", plot=trans_hyp_ret_plot, height=4, width=6)


trans_hyp_ret_log_plot <- ggplot() +
  stat_summary(data=transporters_long, aes(x=Group, y=log2(rge), fill=Tissue),
               fun = 'mean', geom="bar", width=0.75, position="dodge") +
  geom_errorbar(data=trans_sumstats, aes(x=Group, ymin=log2_mean-log2_sd,
                                         ymax=log2_mean+log2_sd, fill=Tissue),
                width=0.2, position=position_dodge(width=0.75)) +
  geom_point(data=transporters_long, aes(x=Group, y=log2(rge), fill=Tissue),
             stat="identity", position=position_dodge(width=0.75)) +
  facet_wrap(~gene, scales="free_y",ncol=1) +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.25))) +
  #scale_fill_manual(values=c("maroon1", "springgreen3")) +
  theme_bw()

trans_hyp_ret_log_plot

#ggsave(filename="../../figures/qRT-PCR/trans_hyp_ret_log.png", plot=trans_hyp_ret_log_plot, height=4, width=6)
#ggsave(filename="../../figures/qRT-PCR/trans_hyp_ret_log.svg", plot=trans_hyp_ret_log_plot, height=4, width=6)
```

### Comparing treatments within each tissue/gene

```{r}
trans_trt_plot <- ggplot() +
  stat_summary(data=transporters_long, aes(x=Group, y=rge, fill=Tissue),
               fun = 'mean', geom="bar", width=0.75, position="dodge") +
  geom_errorbar(data=trans_sumstats, aes(x=Group, ymin=mean-sd, ymax=mean+sd, fill=Tissue),
                width=0.2, position=position_dodge(width=0.75)) +
  geom_point(data=transporters_long, aes(x=Group, y=rge, fill=Tissue),
             stat="identity", position=position_dodge(width=0.75)) +
  facet_grid(gene~Tissue, scales="free_y") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  #scale_fill_manual(values=c("maroon1", "springgreen3")) +
  theme_bw()

trans_trt_plot

#ggsave(filename="../../figures/qRT-PCR/trans_trt.png", plot=trans_trt_plot, height=4, width=6)
#ggsave(filename="../../figures/qRT-PCR/trans_trt.svg", plot=trans_trt_plot, height=4, width=6)


trans_trt_log_plot <- ggplot() +
  stat_summary(data=transporters_long, aes(x=Group, y=log2(rge), fill=Tissue),
               fun = 'mean', geom="bar", width=0.75, position="dodge") +
  geom_errorbar(data=trans_sumstats, aes(x=Group, ymin=log2_mean-log2_sd,
                                         ymax=log2_mean+log2_sd, fill=Tissue),
                width=0.2, position=position_dodge(width=0.75)) +
  geom_point(data=transporters_long, aes(x=Group, y=log2(rge), fill=Tissue),
             stat="identity", position=position_dodge(width=0.75)) +
  facet_grid(gene~Tissue, scales="free_y") +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.25))) +
  #scale_fill_manual(values=c("maroon1", "springgreen3")) +
  theme_bw()

trans_trt_log_plot

#ggsave(filename="../../figures/qRT-PCR/trans_trt_log.png", plot=trans_trt_log_plot, height=4, width=6)
#ggsave(filename="../../figures/qRT-PCR/trans_trt_log.svg", plot=trans_trt_log_plot, height=4, width=6)
```

### Comparing P-gp to Bcrp

```{r}
trans_g2g_plot <- ggplot() +
  stat_summary(data=transporters_long, aes(x=gene, y=rge, fill=gene),
               fun = 'mean', geom="bar", width=0.75, position="dodge") +
  geom_errorbar(data=trans_sumstats, aes(x=gene, ymin=mean-sd,
                                         ymax=mean+sd, fill=gene),
                width=0.2, position=position_dodge(width=0.75)) +
  geom_point(data=transporters_long, aes(x=gene, y=rge, fill=gene),
             position=position_dodge(width=0.75)) +
  facet_grid(Tissue~Group, scales="free_y") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  scale_fill_manual(values=c("springgreen3", "hotpink")) +
  theme_bw()

trans_g2g_plot

#ggsave(filename="../../figures/qRT-PCR/trans_gene_to_gene.png", plot=trans_g2g_plot, height=4, width=6)
#ggsave(filename="../../figures/qRT-PCR/trans_gene_to_gene.svg", plot=trans_g2g_plot, height=4, width=6)


trans_g2g_log_plot <- ggplot() +
  stat_summary(data=transporters_long, aes(x=gene, y=log2(rge), fill=gene),
               fun = 'mean', geom="bar", width=0.75, position="dodge") +
  geom_errorbar(data=trans_sumstats, aes(x=gene, ymin=log2_mean-log2_sd,
                                         ymax=log2_mean+log2_sd, fill=gene),
                width=0.2, position=position_dodge(width=0.75)) +
  geom_point(data=transporters_long, aes(x=gene, y=log2(rge), fill=gene),
             position=position_dodge(width=0.75)) +
  facet_grid(Tissue~Group, scales="free_y") +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.25))) +
  scale_fill_manual(values=c("springgreen3", "hotpink")) +
  theme_bw()

trans_g2g_log_plot

#ggsave(filename="../../figures/qRT-PCR/trans_g2g_log.png", plot=trans_g2g_log_plot, height=4, width=6)
#ggsave(filename="../../figures/qRT-PCR/trans_g2g_log.svg", plot=trans_g2g_log_plot, height=4, width=6)
```

