---
title: "Analysis of Sex Differences in Morphine Deposition in the Plasma, Retina & Brain"
author: "N Bergum & CT Berezin"
date: "7/12/2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(car)
library(tidyverse)
library(ggpubr)
library(rstatix)
library(ggthemes)
library(lme4)
library(EnvStats)
library(lmerTest)
```

## Reading in & tidying the data

```{r}
sex_diff <- read.csv("../data/morphine_retina_cx_plasma_sex_diff.csv",
                     fileEncoding = 'UTF-8-BOM')
sex_diff$conc_wt <- (sex_diff$raw_conc/sex_diff$tissue_weight)
sex_diff$log_conc_wt <- log(sex_diff$conc_wt)
sex_diff$log_conc_raw <- log(sex_diff$raw_conc)

sex_diff <- sex_diff %>% mutate(tissue = factor(as.factor(tissue),
                                                c("plasma", "retina", "brain")),
                                group = factor(as.factor(group),
                                               c("male", "lowE", "highE")),
                                animal = as.factor(animal))

str(sex_diff)

#test for outliers
test_out <- rosnerTest(sex_diff$conc_wt,
  k = 4
)
test_out

SumStat_tissue <- dplyr::summarise(group_by(sex_diff, tissue),
  n = n(),
  mean_tissue_wt = mean(tissue_weight),
  sd_tissue_wt = sd(tissue_weight),
  se_tissue_wt = sd_tissue_wt/sqrt(n),
)
SumStat_tissue

SumStat_all <- dplyr::summarise(group_by(sex_diff, tissue),
  n = n(),
  mean_conc_raw = mean(raw_conc),
  sd_conc_raw = sd(raw_conc),
  se_conc_raw = sd_conc_raw/sqrt(n),
  mean_conc_raw_log = mean(log(raw_conc)),
  sd_conc_raw_log = sd(log(raw_conc)),
  se_conc_raw_log = sd_conc_raw_log/sqrt(n),
  mean_conc_wt = mean(conc_wt),
  sd_conc_wt = sd(conc_wt),
  se_conc_wt = sd_conc_wt/sqrt(n),
  log_conc = mean(log(conc_wt)),
  sd_conc_log = sd(log(conc_wt)),
  se_conc_log = sd_conc_log/sqrt(n),
  mean_wt = mean(body_wt),
  sd_wt = sd(body_wt),
  mean_dose = mean(dose),
  sd_dose = sd(dose),
)

SumStat_all

SumStat_sexdiff <- dplyr::summarise(group_by(sex_diff, tissue, group),
  n = n(),
  mean_conc_raw = mean(raw_conc),
  sd_conc_raw = sd(raw_conc),
  se_conc_raw = sd_conc_raw/sqrt(n),
  mean_conc_raw_log = mean(log(raw_conc)),
  sd_conc_raw_log = sd(log(raw_conc)),
  se_conc_raw_log = sd_conc_raw_log/sqrt(n),
  mean_conc_wt = mean(conc_wt),
  sd_conc_wt = sd(conc_wt),
  se_conc_wt = sd_conc_wt/sqrt(n),
  log_conc = mean(log(conc_wt)),
  sd_conc_log = sd(log(conc_wt)),
  se_conc_log = sd_conc_log/sqrt(n),
  mean_wt = mean(body_wt),
  sd_wt = sd(body_wt),
  mean_dose = mean(dose),
  sd_dose = sd(dose),
)

SumStat_sexdiff
```

## Brain, Retina & Plasma - raw concentrations

Visualizations for potential sex and tissue-dependent differences

```{r}
plotall_tissue <- ggplot() + 
  geom_bar(data=SumStat_sexdiff, aes(x=group, y=mean_conc_raw, fill=tissue),
           stat="identity", position=position_dodge(width=0.75), width=0.75) +
  geom_errorbar(data=SumStat_sexdiff, aes(x=group, ymin=mean_conc_raw - se_conc_raw,
                                      ymax=mean_conc_raw + se_conc_raw, fill=tissue),
                position=position_dodge(width=0.75), width=.25) + 
  geom_point(data=sex_diff, aes(x=group, y=raw_conc, fill=tissue),
             position=position_dodge(width=0.75)) +
  scale_y_continuous(name="[Morphine] (ng/mL)",
                     expand = expansion(mult = c(0, 0.1))) +
  theme_bw(base_size = 12) %+replace%
    theme(legend.title = element_blank()) +
  scale_fill_manual(values=c("red", "dodgerblue", "goldenrod1"),
                    labels=c("plasma" = "Plasma", "retina" = "Retina",
                            "brain" = "Brain")) +
  scale_x_discrete(labels=c("male" = "Males", "lowE" = "Low E/P females",
                            "highE" = "High E/P females")) +
  xlab("")

plotall_tissue
#ggsave("../figures/conc_all_tissues.png", plot=plotall_tissue, width=6, height=4)
#ggsave("../figures/conc_all_tissues.svg", plot=plotall_tissue, width=6, height=4)

plotall_tissueL <- ggplot() + 
  geom_bar(data=SumStat_sexdiff, aes(x=group, y=mean_conc_raw_log, fill=tissue),
           stat="identity", position=position_dodge(width=0.75), width=0.75) +
  geom_errorbar(data=SumStat_sexdiff, aes(x=group,ymin=mean_conc_raw_log - se_conc_raw_log,
                                      ymax=mean_conc_raw_log + se_conc_raw_log, fill=tissue),
                position=position_dodge(width=0.75), width=.25) +
  geom_point(data=sex_diff, aes(x=group, y=log_conc_raw, fill=tissue),
             position=position_dodge(width=0.75)) +
  scale_y_continuous(name="ln([Morphine] (ng/mL))",
                     expand = expansion(mult = c(0, 0.2))) +
  theme_bw(base_size = 12) %+replace%
    theme(legend.title = element_blank()) +
  scale_fill_manual(values=c("red", "dodgerblue", "goldenrod1"),
                    labels=c("plasma" = "Plasma", "retina" = "Retina",
                            "brain" = "Brain")) +
  scale_x_discrete(labels=c("male" = "Males", "lowE" = "Low E/P females",
                            "highE" = "High E/P females")) +
  xlab("")

plotall_tissueL
#ggsave("../figures/conc_all_tissues_log.png", plot=plotall_tissueL, width=6, height=4)
#ggsave("../figures/conc_all_tissues_log.svg", plot=plotall_tissueL, width=6, height=4)

plotall_stage <- ggplot() + 
  geom_bar(data=SumStat_sexdiff, aes(x=tissue, y=mean_conc_raw, fill=group),
           stat="identity", position=position_dodge(width=0.75), width=0.75) +
  geom_errorbar(data=SumStat_sexdiff, aes(x=tissue,ymin=mean_conc_raw - se_conc_raw,
                                      ymax=mean_conc_raw + se_conc_raw, fill=group),
                position=position_dodge(width=0.75), width=.25) + 
  geom_point(data=sex_diff, aes(x=tissue, y=raw_conc, fill=group),
             position=position_dodge(width=0.75)) +
  scale_y_continuous(name="[Morphine] (ng/mL)",
                     expand = expansion(mult = c(0, 0.1))) +
  theme_bw(base_size = 12) %+replace%
    theme(legend.title = element_blank()) +
  scale_fill_manual(labels=c("male" = "Males", "lowE" = "Low E/P females",
                            "highE" = "High E/P females"),
                      values=c("gray50", "darkolivegreen3", "lightpink2")) +
  scale_x_discrete(labels=c("plasma" = "Plasma", "retina" = "Retina",
                            "brain" = "Brain")) +
  xlab("")

plotall_stage
ggsave("../figures/conc_all_stage.png", plot=plotall_stage, width=6, height=4)
ggsave("../figures/conc_all_stage.svg", plot=plotall_stage, width=6, height=4)

plotall_stageL <- ggplot() + 
  geom_bar(data=SumStat_sexdiff, aes(x=tissue, y=mean_conc_raw_log, fill=group),
           stat="identity", position=position_dodge(width=0.75), width=0.75) +
  geom_errorbar(data=SumStat_sexdiff, aes(x=tissue,ymin=mean_conc_raw_log - se_conc_raw_log,
                                      ymax=mean_conc_raw_log + se_conc_raw_log, fill=group),
                position=position_dodge(width=0.75), width=.25) +
  geom_point(data=sex_diff, aes(x=tissue, y=log_conc_raw, fill=group),
             position=position_dodge(width=0.75)) +
  scale_y_continuous(name="ln([Morphine] (ng/mL))",
                     expand = expansion(mult = c(0, 0.2))) +
  theme_bw(base_size = 12) %+replace%
    theme(legend.title = element_blank()) +
  scale_fill_manual(labels=c("male" = "Males", "lowE" = "Low E/P females",
                            "highE" = "High E/P females"),
                      values=c("gray50", "darkolivegreen3", "lightpink2")) +
  scale_x_discrete(labels=c("plasma" = "Plasma", "retina" = "Retina",
                            "brain" = "Brain")) +
  xlab("")

plotall_stageL
#ggsave("../figures/conc_all_stage_log.png", plot=plotall_stageL, width=6, height=4)
#ggsave("../figures/conc_all_stage_log.svg", plot=plotall_stageL, width=6, height=4)
```

Statistical analysis

```{r}
lmer_diff_all <- lmer(raw_conc ~ group*tissue+(1|animal), data =sex_diff)
plot(lmer_diff_all, type=c("p","smooth"), col.line=1)
lattice::qqmath(lmer_diff_all)

lmer_diff_all_log <- lmer(log(raw_conc) ~ group*tissue+(1|animal), data =sex_diff)
plot(lmer_diff_all_log, type=c("p","smooth"), col.line=1)
lattice::qqmath(lmer_diff_all_log)

anova(lmer_diff_all_log)

emmeans::emmeans(lmer_diff_all_log, pairwise ~ tissue| group)$contrasts
emmeans::emmeans(lmer_diff_all_log, pairwise ~ group| tissue)$contrasts



#shapiro.test(resid(lmer_diff))
```

## Brain & Retina normalized by tissue weight

```{r}
sex_diff <- sex_diff %>% filter(tissue != "plasma")
SumStat_sexdiff <- SumStat_sexdiff %>% filter(tissue != "plasma")
SumStat_sexdiff

plot_tissue <- ggplot() + 
  geom_bar(data=SumStat_sexdiff, aes(x=group, y=mean_conc_wt, fill=tissue),
           stat="identity", position=position_dodge(width=0.75), width=0.75) +
  geom_errorbar(data=SumStat_sexdiff, aes(x=group, ymin=mean_conc_wt - se_conc_wt,
                                      ymax=mean_conc_wt + se_conc_wt, fill=tissue),
                position=position_dodge(width=0.75), width=.25) + 
  geom_point(data=sex_diff, aes(x=group, y=conc_wt, fill=tissue),
             position=position_dodge(width=0.75)) +
  scale_y_continuous(name="[Morphine] per mg tissue (ng/mg)",
                     expand = expansion(mult = c(0, 0.1))) +
  theme_bw(12) %+replace%
    theme(legend.title = element_blank()) +
  scale_fill_manual(values=c("dodgerblue", "goldenrod1"),
                    labels=c("retina" = "Retina", "brain" = "Brain")) +
  scale_x_discrete(labels=c("male" = "Males", "lowE" = "Low E/P females",
                            "highE" = "High E/P females")) +
  xlab("")

plot_tissue
#ggsave("../figures/conc_tissue.png", plot=plot_tissue, width=6, height=4)
#ggsave("../figures/conc_tissue.svg", plot=plot_tissue, width=6, height=4)

plot_tissueL <- ggplot() + 
  geom_bar(data=SumStat_sexdiff, aes(x=group, y=log_conc, fill=tissue),
           stat="identity", position=position_dodge(width=0.75), width=0.75) +
  geom_errorbar(data=SumStat_sexdiff, aes(x=group,ymin=log_conc - se_conc_log,
                                      ymax=log_conc + se_conc_log, fill=tissue),
                position=position_dodge(width=0.75), width=.25) +
  geom_point(data=sex_diff, aes(x=group, y=log_conc_wt, fill=tissue),
             position=position_dodge(width=0.75)) +
  scale_y_continuous(name="ln([Morphine] per mg tissue (ng/mg))",
                     expand = expansion(mult = c(0, 0.2))) +
  theme_bw(12) %+replace%
    theme(legend.title = element_blank()) +
  scale_fill_manual(values=c("dodgerblue", "goldenrod1"),
                    labels=c("retina" = "Retina", "brain" = "Brain")) +
  scale_x_discrete(labels=c("male" = "Males", "lowE" = "Low E/P females",
                            "highE" = "High E/P females")) +
  xlab("")

plot_tissueL
ggsave("../figures/conc_tissue_log.png", plot=plot_tissueL, width=6, height=4)
ggsave("../figures/conc_tissue_log.svg", plot=plot_tissueL, width=6, height=4)

plot_stage <- ggplot() + 
  geom_bar(data=SumStat_sexdiff, aes(x=tissue, y=mean_conc_wt, fill=group),
           stat="identity", position=position_dodge(width=0.75), width=0.75) +
  geom_errorbar(data=SumStat_sexdiff, aes(x=tissue,ymin=mean_conc_wt - se_conc_wt,
                                      ymax=mean_conc_wt + se_conc_wt, fill=group),
                position=position_dodge(width=0.75), width=.25) + 
  geom_point(data=sex_diff, aes(x=tissue, y=conc_wt, fill=group),
             position=position_dodge(width=0.75)) +
  scale_y_continuous(name="[Morphine] per mg tissue (ng/mg)",
                     expand = expansion(mult = c(0, 0.1))) +
  theme_bw(base_size = 12) %+replace%
    theme(legend.title = element_blank(),
          legend.spacing.y = unit(0.1, 'cm')) +
  ## next line needed to make previous line work
  guides(fill = guide_legend(byrow = TRUE)) +
  scale_fill_manual(values=c("gray50", "darkolivegreen3", "lightpink2"),
                    labels=c("male" = "Males", "lowE" = "Low E/P\nfemales",
                            "highE" = "High E/P\nfemales")) +
  scale_x_discrete(labels=c("retina" = "Retina", "brain" = "Brain")) +
  xlab("")

plot_stage
#ggsave("../figures/conc_stage.png", plot=plot_stage, width=6, height=4)
#ggsave("../figures/conc_stage.svg", plot=plot_stage, width=6, height=4)


plot_stageL <- ggplot() + 
  geom_bar(data=SumStat_sexdiff, aes(x=tissue, y=log_conc, fill=group),
           stat="identity", position=position_dodge(width=0.75), width=0.75) +
  geom_errorbar(data=SumStat_sexdiff, aes(x=tissue,ymin=log_conc - se_conc_log,
                                          ymax=log_conc + se_conc_log, fill=group),
                position=position_dodge(width=0.75), width=.25) +
  geom_point(data=sex_diff, aes(x=tissue, y=log_conc_wt, fill=group),
             position=position_dodge(width=0.75)) +
  scale_y_continuous(name="ln([Morphine] per mg tissue (ng/mg))",
                     expand = expansion(mult = c(0, 0.2))) +
  theme_bw(base_size = 12) %+replace%
    theme(legend.title = element_blank(),
          legend.spacing.y = unit(0.1, 'cm')) +
  ## next line needed to make previous line work
  guides(fill = guide_legend(byrow = TRUE)) +
  scale_fill_manual(values=c("gray50", "darkolivegreen3", "lightpink2"),
                    labels=c("male" = "Males", "lowE" = "Low E/P\nfemales",
                            "highE" = "High E/P\nfemales")) +
  scale_x_discrete(labels=c("retina" = "Retina", "brain" = "Brain")) +
  xlab("")

plot_stageL
#ggsave("../figures/conc_stage_log.png", plot=plot_stageL, width=6, height=4)
#ggsave("../figures/conc_stage_log.svg", plot=plot_stageL, width=6, height=4)
```

Statstical analysis for differences between groups and tissues

```{r}
lmer_diff <- lmer(log(conc_wt) ~ group*tissue+(1|animal), data =sex_diff)
plot(lmer_diff, type=c("p","smooth"), col.line=1)
lattice::qqmath(lmer_diff)

anova(lmer_diff)
emmeans::emmeans(lmer_diff, pairwise ~ tissue| group)$contrasts
emmeans::emmeans(lmer_diff, pairwise ~ group| tissue)$contrasts
#shapiro.test(resid(lmer_diff))
```

## Sex differences analysis exclusively for animals that were littermates

```{r}
litter <- subset(sex_diff,  animal %in% c("1M","2M","3M","4M","5M","6M","7M","8M","9M","18F","19F","20F","21F","22F","24F","25F","26F"))
litter
lmer_litt <- lmer(log(conc_wt) ~ group*tissue+(1|animal), data =litter)
plot(lmer_litt, type=c("p","smooth"), col.line=1)
lattice::qqmath(lmer_litt)
anova(lmer_litt)
emmeans::emmeans(lmer_litt, pairwise ~ tissue| group)
emmeans::emmeans(lmer_litt, pairwise ~ group| tissue)
shapiro.test(resid(lmer_litt))
```

Visualizations

```{r}
SumStat_sexlitt <- dplyr::summarise(group_by(litter, tissue, group),
n = n(),
mean_conc_wt = mean(conc_wt),
sd_conc_wt = sd(conc_wt),
se_conc_wt = sd_conc_wt/sqrt(n),
log_conc = mean(log(conc_wt)),
sd_conc_log = sd(log(conc_wt)),
se_conc_log = sd_conc_log/sqrt(n),
mean_wt = mean(body_wt),
sd_wt = sd(body_wt),
mean_dose = mean(dose),
sd_dose = sd(dose),
)

SumStat_sexlitt

plot_tissue_litt <- ggplot() + 
  geom_bar(data=SumStat_sexlitt, aes(x=group, y=mean_conc_wt, fill=tissue), stat="identity", position=position_dodge(width=0.5), width=0.5) +
  geom_errorbar(data=SumStat_sexlitt, aes(x=group,ymin=mean_conc_wt - se_conc_wt, ymax=mean_conc_wt + se_conc_wt, fill=tissue), position=position_dodge(width=0.5), width=.2,) + 
  geom_point(data=litter, aes(x=group, y=conc_wt, fill=tissue), position=position_dodge(width=0.5)) +
  scale_y_continuous(name="Morphine concentration per mg tissue (ng/mg)") + theme_bw()
plot_tissue_litt

plot_stage_litt <- ggplot() + 
  geom_bar(data=SumStat_sexlitt, aes(x=tissue, y=mean_conc_wt, fill=group), stat="identity", position=position_dodge(width=0.5), width=0.5) +
  geom_errorbar(data=SumStat_sexlitt, aes(x=tissue,ymin=mean_conc_wt - se_conc_wt, ymax=mean_conc_wt + se_conc_wt, fill=group), position=position_dodge(width=0.5), width=.2,) + 
  geom_point(data=litter, aes(x=tissue, y=conc_wt, fill=group), position=position_dodge(width=0.5)) +
  scale_y_continuous(name="Morphine concentration per mg tissue (ng/mg)") + theme_bw()
plot_stage_litt

plot_tissueL_litt <- ggplot() + 
  geom_bar(data=SumStat_sexlitt, aes(x=group, y=log_conc, fill=tissue), stat="identity", position=position_dodge(width=0.5), width=0.5) +
  geom_errorbar(data=SumStat_sexlitt, aes(x=group,ymin=log_conc - se_conc_log, ymax=log_conc + se_conc_log, fill=tissue), position=position_dodge(width=0.5), width=.2,) +  geom_point(data=litter, aes(x=group, y=log_conc_wt, fill=tissue), position=position_dodge(width=0.5)) +
  scale_y_continuous(name="ln(Morphine concentration per mg tissue (ng/mg))") + theme_bw()
plot_tissueL_litt

plot_stageL_litt <- ggplot() + 
  geom_bar(data=SumStat_sexlitt, aes(x=tissue, y=log_conc, fill=group), stat="identity", position=position_dodge(width=0.5), width=0.5) +
  geom_errorbar(data=SumStat_sexlitt, aes(x=tissue,ymin=log_conc - se_conc_log, ymax=log_conc + se_conc_log, fill=group), position=position_dodge(width=0.5), width=.2,) +  geom_point(data=litter, aes(x=tissue, y=log_conc_wt, fill=group), position=position_dodge(width=0.5)) +
  scale_y_continuous(name="ln(Morphine concentration per mg tissue (ng/mg))") + theme_bw()
plot_stageL_litt
```


## Comparisons between estrus stages


```{r}
SumStat_sexstage <- dplyr::summarise(group_by(sex_diff, tissue, stage),
n = n(),
mean_conc_wt = mean(conc_wt),
sd_conc_wt = sd(conc_wt),
se_conc_wt = sd_conc_wt/sqrt(n),
log_conc = mean(log(conc_wt)),
sd_conc_log = sd(log(conc_wt)),
se_conc_log = sd_conc_log/sqrt(n),
mean_wt = mean(body_wt),
sd_wt = sd(body_wt),
mean_dose = mean(dose),
sd_dose = sd(dose),
)

plot_stage_tissue <- ggplot() + 
  geom_bar(data=SumStat_sexstage, aes(x=stage, y=mean_conc_wt, fill=tissue), stat="identity", position=position_dodge(width=0.5), width=0.5) +
  geom_errorbar(data=SumStat_sexstage, aes(x=stage,ymin=mean_conc_wt - se_conc_wt, ymax=mean_conc_wt + se_conc_wt, fill=tissue), position=position_dodge(width=0.5), width=.2,) + 
  geom_point(data=sex_diff, aes(x=stage, y=conc_wt, fill=tissue), position=position_dodge(width=0.5)) +
  scale_y_continuous(name="Morphine concentration per mg tissue (ng/mg)") + theme_bw()
plot_stage_tissue

plot_stages <- ggplot() + 
  geom_bar(data=SumStat_sexstage, aes(x=tissue, y=mean_conc_wt, fill=stage), stat="identity", position=position_dodge(width=0.5), width=0.5) +
  geom_errorbar(data=SumStat_sexstage, aes(x=tissue,ymin=mean_conc_wt - se_conc_wt, ymax=mean_conc_wt + se_conc_wt, fill=stage), position=position_dodge(width=0.5), width=.2,) + 
  geom_point(data=sex_diff, aes(x=tissue, y=conc_wt, fill=stage), position=position_dodge(width=0.5)) +
  scale_y_continuous(name="Morphine concentration per mg tissue (ng/mg)") + theme_bw()
plot_stages

lmer_stage <- lmer(log(conc_wt) ~ stage*tissue+(1|animal), data =sex_diff)
plot(lmer_stage, type=c("p","smooth"), col.line=1)
lattice::qqmath(lmer_stage)
anova(lmer_stage)
emmeans::emmeans(lmer_stage, pairwise ~ tissue| stage)
emmeans::emmeans(lmer_stage, pairwise ~ stage| tissue)
shapiro.test(resid(lmer_stage))
```